# Transformations and Expectations {#transformations}

```{r include = FALSE}
library(ggplot2)
```


## Chapter Notes


## Questions

### Ex 2.1 {-}

**Question**

> In each of the following find the pdf of $Y$. Show that the pdf integrates to 1. 
>
> (a) $Y = X^3$ and $f_X(x) = 42x^5(1-x)$,  $0 < x < 1$
>
> (b) $Y = 4X +3$ and $f_X(x) = 7e^{-7x}$,  $0 < x < \infty$
>
> (c) $Y = X^2$ and $f_X(x) = 30x^2(1-x)^2$,  $0 < x < 1$

**Answer**

Will need Theorem 2.1.5 on page 51/77.

(a) 

Our transformation $Y=X^3$ is monotone on $0 < x < 1$ 

```{r echo = FALSE}
ggplot()+
  geom_function(fun = ~ .x^3)+
  xlim(0,1)+
  xlab("x")
```

and so by Theorem 2.1.5 our pdf is:

$$
\begin{aligned}
f_Y(y) &= f_X(y^{1/3}) |\frac{d}{dy}y^{1/3}| \\
& = 42(y^{1/3})^5(1-y^{1/3})\frac{1}{3}y^{\left(-2/3\right)}\\
&=
\end{aligned}
$$

### Ex 2.3 {-}

**Question**

> Suppose $X$ has the geometric pmf $f_X(x) = \frac{1}{3}(\frac{2}{3})^x$, $x = 0, 1, 2, \dots$. Determine the probability distribution of $Y = X/(X + 1)$. 
Note that here both $X$ and $Y$ are discrete random variables. To specify the probability distribution of $Y$, specify its pmf.

**Answer**


### Ex 2.6 {-}

**Question**

> In each of the following find the pdf of $Y$ and show that the pdf integrates to 1.
>
> (a) $f_X(x) = \frac{1}{2}e^{-|x|}$, $-\infty < x < \infty$; $Y = |X|^3$
>
> (b) $f_X(x) = \frac{3}{8} (x+1)^2$, $-1 < x < 1$; $Y = 1-X^2$
>
> (c) $f_X(x) = \frac{1}{2} (x+1)^2$, $-1 < x < 1$; $Y = 1-X^2$ if $X\leq 0$ and $Y=1-X$ if $X>0$
 

**Answer**

## Further Reading {-}
